---
title: "Lecture 7"
author: "Michal Kubi&#353;ta"
date: "26 February 2018"
output:
  ioslides_presentation:
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, ffmpeg.format = "mp4",
                      cache = TRUE, fig.align = 'center',
                      fig.width = 8, fig.height = 5)
library(magrittr)
library(ggplot2)
library(rpart)
library(rattle)
library(gridExtra)
```

## Structure
1.) Decission trees
2.) Customer churn
3.) Boosting and Bagging
4.) Random forests

# Decission trees

## Introduction to decission rules
- if-else conditions used to split data
- aimed at increasing "purity"
    - InfoGain / Gini
    - decrease of variation
- if build hierarchically, they form a tree

## Trees
- classification & regression
- very greedy
    - always optimising on one variable only!
- more flexible than linear models
    - can easily handle non-linear relations
    - automatic variable selection
- can't predict out of scope of the input data
    - can't be used for time series (on their own)

## Tree types
- rpart
    - recursive partitioning
    - both classification / regression
    - Gini 
- C.5
    - only classification
    - InfoGain
    
## Example trees

<div class="columns-2">
```{r tree_example_num, fig.align='left', fig.width=4}
rpart(gear ~ ., mtcars) %>% fancyRpartPlot()
```

```{r tree_example_fac, fig.align='right', fig.width=4}
rpart(as.factor(gear) ~ ., mtcars) %>% fancyRpartPlot()
```
</div>

## Summary
- rpart::rpart
- C5O::C5.0.default
- handle complex relations well
- automatic variable selection
    - good for large datasets
- very greedy
- unstable
    - bad for small datasets

# Customer churn

## What is churn?
- customer attrition
- loss of clients
- targetted mostly by
    - banks, insurance companies
    - telcom, intenet, TV providers
- mostly service companies
- cheeper to retain customers than to find new ones

## Much ado churn?
- Why do those companies care so much?
    - customers ~ free money
- financial institutions
    - you keep your customers money
- communication & TV
    - customer pays every month
- service providers with high initial (fixed) costs and very low variable costs
    - how much are the actuall costs of transfering 1 GB of data?
    - how much are the costs of providing insurance to customer?

# Boosting & Bagging

## Boosting
- ensembling a large number of week learners (models) into one strong
- linear
    - first model
    - second is trained on residuals of the first
    - third is trained on the residuals (of residuals) of the second model
- subsetting observations
    - more weight to missclasiffied ones
- increasing accuracy
    - prone to overfitting

## Bagging
- ensembling a large number of week learners (models) into one strong
- in parallel
    - independent models
- subsetting observations and variables
    - observations sampled with repetitions
        - one original observation can be used more than one time in each of the models
- increasing stability

# Random forests

## Welcome to the forest
- bagging algorithm
- build from trees
- good accuracy and stability

>- black box

## Opening the black box
- var imp
- forestfloor
- predictions

