---
title: "Lecture 5"
author: "Michal Kubi&#353;ta"
date: "5 February 2018"
output:
  ioslides_presentation:
    widescreen: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, ffmpeg.format = "mp4",
                      cache = TRUE)
library(magrittr)
library(ggplot2)
library(gganimate)
```

## Structure
1. Regression
2. PCA

# Regression

##  {.flexbox .vcenter}

<font size="24" color="orange"> **You tell me...**  
recall you statistics / econometrics class </font>

## Econometric approach
OLS

- statistical method
- assumptions (Gauss-Markov)
    1. linear relationship
        - transformations
    2. zero expected error ($E(\epsilon_{i}) = 0$)
        - intercept
    3. no perfect multicollinearity (full X matrix rank)
    4. <font color="orange">**homoskedasticity**</font> ($E(\epsilon_{i}^{2}) = const$)
        - slope
        
## OLS estimation
$$\hat\beta = (X^{T}X)^{-1}X^{T}y$$
**n** observations, **k** variables (\* = OK for small **k**)

- $X^T \; complextity \; O(nk)$ *
- $X^{T}X = T_{k*k} \; complexity \; O(n^{2})$ <font color="red">**!**</font>
- $T^{-1} = U_{k*k} \; complexity \; O(k^{3})$ *
- $UX^{T} = V_{k*n} \; complexity \; O(k^{2}n)$ *
- $Vy = W_{k*1} \; complexity \; O(kn)$ *

>- what if **k** is not small?


## Setup the estimation of the linear relationship
 how to uniquely describe a line (2D) ?
 
>- 2 parameters, slope and a point
>- there is only 1 parameter to estimate - why?
>- $(\bar X,\bar Y)$
>- optimisation task

## (i) random guess

```{r rndAnim, fig.show='animate'}
rand <- list()

xhat <- mean(mtcars$drat)
yhat <- mean(mtcars$wt)
xmin <- min(mtcars$drat)
xmax <- max(mtcars$drat)

for(i in 1:10){
  slope <- runif(1,-10,10)
  ymax <- yhat + slope*(xmax - xhat)
  ymin <- yhat + slope*(xmin - xhat)
  rand[[i]] <- rbind(c(xmin, ymin, i),c(xmax, ymax, i))
}

rand <- do.call(rbind.data.frame, rand)
colnames(rand) <- c("x","y","iter")

rndPlot <-
  ggplot(rand, aes(x = x, y = y, frame = iter))+
  geom_line()+
  geom_point(data = mtcars, aes(x = drat, y = wt), inherit.aes = F)

gganimate(rndPlot)
```

## (ii) random guess + evolution

- save the parameters and SSR of first iteration - best estimation
- if other parameters lead to lower SSR
    - rewrite the best estimation
- new estimation close to the parameters of the best estimation 
- <font color="orange">**very simple**</font> MCMC process
  
## (iii) random guess + evolution + reproduction

1. generate many random "estimates" of slope
2. find the best individuals
    - less then threshold SSR
    - or given number of best-fit individuals
3. let them breed
    - averaging, linear combinations
4. find the best individuals of the extended population
5. repeat 3. and 4. until convergence / stopping condition
- simple evolution algorithm

## Evolution estimation

```{r evoAnim, fig.show='animate'}
xhat <- mean(mtcars$drat)
yhat <- mean(mtcars$wt)
xmin <- rep(min(mtcars$drat), 4)
xmax <- rep(max(mtcars$drat), 4)

evo <- list()

for(i in 1:10){
  if(i == 1){
    slope <- runif(10,-10,10)
  } else{
    slope <- c(slope, (slope[1]+slope[-1])/2)
  }
  
  SSR <- c()
  for(j in slope){
    SSR <- c(SSR, sum((mtcars$wt - (mtcars$drat - xhat) * j - yhat)^2))
  }
  
  parents <- order(SSR)[1:4]
  
  slope <- slope[parents]
  
  ymax <- yhat + slope*(xmax - xhat)
  ymin <- yhat + slope*(xmin - xhat)
  iter <- i
  col <- rep(1:4, each = 2)
  
  evo[[i]] <- cbind(c(rbind(xmin,xmax)), c(rbind(ymin, ymax)), iter, col)
  
}
evo <- do.call(rbind.data.frame, evo)
colnames(evo) <- c("x", "y", "iter", "col")


evoPlot <-
  ggplot(evo, aes(x = x, y = y, frame = iter, col = as.factor(col)))+
  geom_line()+
  geom_point(data = mtcars, aes(x = drat, y = wt), inherit.aes = F)+
  theme(legend.position = 'none')

gganimate(evoPlot)
```

## (iv) gradient descent
1. calculate the gradient of the function
    - $SSR = \sum_i (y_{i} - \hat y_{i})^{2} =
    \sum_i (y_{i} - (a + bx_{i}))^{2} =$  
    $\sum_i (y_{i} - (\hat y - b \hat x) - bx_{i})^{2} = \sum_i (y_{i} - \hat y - b(x_{i} - \hat x))^{2}$
    - $\frac{\nabla SSR}{n} = \frac{\partial SSR}{\partial b} * \frac{1}{n} = H(b)$
2. generate initial random "estimate" of slope = $b_{0}$
3. move a little in the direction of the gradient
    - $b_{n} = b_{n-1} - \alpha H(b)$
    - $\alpha = parameter$
4. repeat 3. until convergence / stopping condition

## Differences
- for the purpose of fitting the linear relationship (no inference)
    - no assumptions
- for inference (prediction)
    - mostly care about homoskedasticity & outliers
       - different error function
    - cross-validation
- watch out for local optimas!
    - ABC optimisation
    
# PCA
